RL-Musician: A Configurable Open-Source Tool for Species Counterpoint Composition

### 2020-05-17 docs/paper/paper.tex
docs: Revise the draft of the paper

The word "species" in the name of the framework indicates that there are some types -- each of them has its own rhythmic pattern. Here, fifth species counterpoint is covered. This type is the highest one and the less restrictive one. Like in other species, there is a given melodic line consisting entirely of whole (semibreve) notes. This line is called cantus firmus. Again, like in other species, a new melodic line of the same duration is going to be created (so called counterpoint line) and these two lines should sound aesthetically pleasant when played simultaneously. The difference between fifth species and other species is that a counterpoint line may change its rhythmic pattern: half notes, quarter notes, eights notes, and suspended whole notes can be used.

@@

To start with, define representation of a musical piece. Suppose that a diatonic scale (like C-major or A-minor) is chosen. Let a line be a list of $(\mathrm{pitch}, \mathrm{duration})$ pairs where $\mathrm{pitch}$ must be from the scale (i.e., only 7 out of 12 notes per octave can be used and so chromaticism is prohibited) and $\mathrm{duration} \in \{1, 2, 4, 8\}$ (i.e., it is measured in eights of measure). In this study, pieces are written exactly in 2 melodic lines (parts, voices) where one of them is cantus firmus and the other one is counterpoint. Definition of cantus firmus implies that every its element has duration equal to 8 and duration of a piece in measures is equal to length of cantus firmus. Denote this length by $m$, where $m$ is usually between 8 and 24. From music point of view, this means that a piece to be created is rather a phrase that can be inserted into a longer composition. Such limitation goes back to species counterpoint, because it is designed exactly for writing short phrases.

Alternatively, a piece can be represented by a piano roll which is a matrix $P$ of size $n \times 8m$ where $n$ is size of available pitches range (hence, $n \le 88$, since there are 88 piano keys) and $8m$ is number of eights of measure in a piece lasting $m$ measures. Matrix element $P_{ij}$ is 1 if $i$-th pitch is sounding at $j$-th eight and 0 otherwise.

@@

Initially, cantus firmus, start note of counterpoint line (occupying the whole first measure except opening pause), end note of counterpoint line (occupying the whole last measure), and range of pitches available for usage in counterpoint line, are set externally. These values are the same for all episodes of Monte Carlo Beam Search. During each episode, a gap between the first note and the last note is reduced by one new note added to the left. This is done until the counterpoint line is finished.

@@

Below figure shows a piece produced given this setup. As Table \ref{table:scores} shows, the piece is close to theoretical optimum.

>>>>

The word "species" in the name of the framework indicates that there are some types -- each of them has its own rhythmic pattern. Here, fifth species counterpoint is covered. This type is the highest one and the less restrictive one. Like in other species, there is a given melodic line consisting entirely of whole (semibreve) notes. This line is called cantus firmus. Again, like in other species, a new melodic line of the same duration is going to be created (so called counterpoint line) and these two lines should sound aesthetically pleasant when played simultaneously. The difference between fifth species and other species is that a counterpoint line may change its rhythmic pattern: half notes, quarter notes, eighths notes, and suspended whole notes can be used.

@@

To start with, define representation of a musical piece. Suppose that a diatonic scale (like C-major or A-minor) is chosen. Let a line be a list of $(\mathrm{pitch}, \mathrm{duration})$ pairs where $\mathrm{pitch}$ must be from the scale (i.e., only 7 out of 12 notes per octave can be used and so chromaticism is prohibited) and $\mathrm{duration} \in \{1, 2, 4, 8\}$ (i.e., it is measured in eighths of measure). In this study, pieces are written exactly in 2 melodic lines (parts, voices) where one of them is cantus firmus and the other one is counterpoint. Definition of cantus firmus implies that every its element has duration equal to 8 and duration of a piece in measures is equal to length of cantus firmus. Denote this length by $m$, where $m$ is usually between 8 and 24. From music point of view, this means that a piece to be created is rather a phrase that can be inserted into a longer composition. Such limitation goes back to species counterpoint, because it is designed exactly for writing short phrases.

Alternatively, a piece can be represented by a piano roll which is a matrix $P$ of size $n \times 8m$ where $n$ is size of available pitches range (hence, $n \le 88$, since there are 88 piano keys) and $8m$ is number of eighths of measure in a piece lasting $m$ measures. Matrix element $P_{ij}$ is 1 if $i$-th pitch is sounding at $j$-th eighth and 0 otherwise.

@@

Initially, cantus firmus, start note of counterpoint line (occupying the whole first measure except opening pause), end note of counterpoint line (occupying the whole last measure), and range of pitches available for usage in counterpoint line, are set externally. These values are the same for all episodes of Monte Carlo Beam Search. During each episode, a gap between the first note and the last note is reduced by one new note added to the left and this continues until the counterpoint line is finished.

@@

Below figure demonstrates a piece produced with this setup. As Table \ref{table:scores} shows, the piece is close to theoretical optimum.
